EDIpack 2.0: A generic and interoperable Lanczos-based solver for quantum impurity problems

### 2025-05-20 01_intro.tex
Updates from Overleaf

Quantum impurity models play a central role in the study of strongly correlated electron systems, capturing the coupling of a small number of localized degrees of freedom with extended non-interacting environment.   
Originally introduced to describe magnetic impurities in metallic hosts, i.e. Kondo problem, these systems progressively became powerful tools for exploring a wide range of quantum many-body phenomena. More recently, their importance extended beyond the realm of condensed matter, influencing understanding of other areas such as quantum information or ultra-cod atomic systems.

At the heart of quantum impurity models is the interaction between a localized impurity, possibly including lattice, orbital and spin states, and a surrounding bath of itinerant particles, usually represented as a conduction band of electrons. 
This setup capture many essential aspects of strong correlation including screening, decoherence and entanglement in a computationally tractable way. 
Most notably, quantum impurity problems lie at the core of dynamical mean-field theory (DMFT) where the complex behavior of the lattice systems is reduced to an effective self-consistent quantum impurity problem.  
By retaining essential local correlations, the DMFT enabled a successful effective description of the most important physical aspects of correlated materials, including Mott insulators, heavy fermion compounds, and high-temperature superconductors.

However, solving the quantum impurity problems emerging in the DMFT context accurately and efficiently remains a
critical challenge, particularly for multi-orbital systems, systems
at low temperatures or endowed with lower symmetry. During many years of development, different advanced numerical techniques have been developed together with cutting edge approaches aimed to reduce computational load.

%EDIPACK 2.0
In this work we introduce  EDIpack 2.0: a flexible, high-performance numerical library designed to address these challenges, providing an exact diagonalization (ED) solver for generic Quantum Impurity Problems. 
The library builds on the foundations of
its predecessor, featuring massively parallel Lanczos-based algorithms while including several significant extensions and 
improvements. The library now supports both zero and finite temperature calculations for different symmetries of the problem within a unified framework, making it suitable for a wide range of physical phenomena, from spin-orbit coupling effect in quantum materials to multi-orbital superconductivity. It also provides direct
access to well-approximated dynamical correlation functions across the
entire complex plane, thus enabling calculations of spectral functions, local susceptibilities, and
other response properties.

@@

%\paragraph{Interoperability}\\
With the growing importance of  quantum impurity models in the study of correlated materials, ensuring compatibility among different methods and developments has become a key challenge.
In this context, interoperability became a core design principle of this new updated version of \NAME. The library is implemented using modern Fortran constructs, with comprehensive C-bindings that enable integration with other programming languages. 
In particular, \NAME includes native C and C++
interface layer which enable development of third-party
extensions. A significant contribution is realized via the Python API, EDIpy, which provides a high-level interface for defining and solving impurity problems. The
Python bindings also enable integration with established scientific
libraries, such as NumPy and SciPy, facilitating data analysis and
rapid prototyping.

A key development feature in \NAME, unlocked by the introduction of Python and C/C++ API, is the direct  interfaces to the TRIQS
(Toolbox for Research on Interacting Quantum Systems)  as
well as to the W2Dynamics software, which are two of the most important and used computational tools in the field. 
This integration provides \NAME users with access to the
extensive data handling and analysis tools in the TRIQS ecosystem,
while also allowing TRIQS and W2Dynamics users to benefit from the
flexibility and performance of \NAME solver. This dual functionality
extends the capabilities of these platforms, supporting
reproducibility and cross-validation of numerical results, and sets a
new standard for interoperability in quantum impurity solvers.

>>>>

Quantum impurity models play a central role in the study of strongly correlated electron systems, capturing the coupling of a small number of localized degrees of freedom with extended non-interacting environment.   
Originally introduced to describe magnetic impurities in metallic hosts and leading to the solution of the Kondo problem, these systems progressively became powerful tools for exploring a wide range of quantum many-body phenomena. More recently, their importance has extended beyond the realm of condensed matter, influencing understanding of other areas such as quantum information or ultra-cold atomic systems.

At the heart of quantum impurity models is the interaction between a localized ``impurity"  and a surrounding bath of itinerant particles, usually represented as a conduction band of electrons. The definition of the impurity is rather generic, and it can contain lattice (i.e. inequivalent sites), orbital and spin local degrees of freedom.
This setup captures many essential aspects of strong correlation including screening, decoherence and entanglement in a computationally tractable way. 
Most notably, quantum impurity problems lie at the core of dynamical mean-field theory (DMFT) where the complex behavior of the lattice systems is reduced to an effective self-consistent quantum impurity problem.  
By accounting for local quantum fluctuations, DMFT can successfully describe of the most important physical features of correlated materials, including Mott insulators, heavy fermion compounds, and high-temperature superconductors.

Accurately and efficiently solving the DMFT quantum impurity problem remains a
critical challenge, particularly for multi-orbital systems, systems
at low temperatures or endowed with lower symmetry. During many years of development, different advanced numerical techniques have been developed together with cutting edge approaches aimed to reduce computational load.

%EDIPACK 2.0
In this work we introduce  \NAME{}: a flexible, high-performance numerical library designed to address these challenges, providing an exact diagonalization (ED) solver for generic Quantum Impurity Problems. 
The library builds on the foundations of
its predecessor, featuring massively parallel Lanczos-based algorithms while including several significant extensions and 
improvements. The library now supports, within a unified framework, both zero and finite temperature calculations for a wide set of impurity models, making it capable of addressing a large set of problems, from spin-orbit coupling effect in quantum materials to multi-orbital superconductivity. It also provides direct
access to well-approximated dynamical correlation functions across the
entire complex plane, thus enabling calculations of spectral functions, local susceptibilities, and other response properties.

@@

%\paragraph{Interoperability}\\
With the growing importance of  quantum impurity models in the study of correlated materials,
standardizazion and interoperability between solution software suites has
become a critical necessity.
\NAME has been expressly designed following this guiding principle. 
While the library itself is implemented using modern Fortran constructs, it also provides an extensive set of C-bindings that enable integration with other programming languages,
be they compiled (C, C++), JIT-compiled (Julia) or interpreted (python). 
A significant contribution is realized via the Python API, EDIpy, which provides a high-level interface for defining and solving impurity problems. The
Python bindings also enable integration with established scientific
libraries, such as NumPy and SciPy, facilitating data analysis and
rapid prototyping.

A key development feature in \NAME, unlocked by the introduction of Python and C/C++ API, 
is the direct  interfaces to the TRIQS (Toolbox for Research on Interacting Quantum Systems)  as
well as to the W2Dynamics software, which are two of the most 
important and used computational tools in the field. \NAME seamlessly slots in as an alternative
solver to the native Quantum Monte Carlo implementations, greatly extending the parameter (and 
especially temperature) range reachable by the simulation while minimizing code production from 
the end user.
At the same time, this integration provides \NAME users with access to the
extensive data handling and analysis tools in the TRIQS ecosystem. 
Overall, interoperability extends the cumulative reach and trustworthiness of all these platforms,
enhancing reproducibility and cross-validation of numerical results, with the ultimate aim of achieving standardization for quantum impurity solvers.
